---
title: 'Discussion 4: Estimating Median and Median Residual Survival Times'
author: "Serge Aleshin-Guendel"
date: "2/1/2021"
output:
  beamer_presentation: default
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives
Using the Kaplan-Meier estimate of the survival function we will focus on:

- How to estimate the median survival time 
- How to estimate the median residual survival time

## Slides

- Slides are available on Canvas
- RMarkdown can be checked out via GitHub

```{sh, eval=F}
git clone
 https://github.com/mbannick/survival-discussion-section.git
```

## Reading in the TRACE data

We will use the TRACE data as in the second discussion section. Note that in 
order to source \texttt{getmedianres.R} as we do below, it needs to be located 
in your working directory

```{r, warning = FALSE}
# Load relevant packages 
library(foreign)
library(survival)

# Load in getmedianres function
source("getmedianres.R")

# Read in TRACE data
trace <- read.dta("traceST11.dta")

# Create a Surv object using the TRACE data
s.trace <- with(trace, Surv(time, status == 9)) 
```

## Estimating the survival function nonparametrically

We will focus today on estimating the median and median residual survival times,
based on Kaplan-Meier estimates of the survival function

\footnotesize
```{r, eval = FALSE}
survfit.trace <- survfit(s.trace ~ 1, data = trace, conf.type = "log-log")
plot(survfit.trace, conf.int = TRUE, xlab = "Time (Years)", 
     ylab = "Survival Probability")
```
\normalsize

## Estimating the survival function nonparametrically

```{r, fig.width = 10, fig.height = 8, echo=FALSE}
survfit.trace <- survfit(s.trace ~ 1, data = trace, conf.type = "log-log")
plot(survfit.trace, conf.int = TRUE, xlab = "Time (Years)", 
     ylab = "Survival Probability")
```


## What is the median survival time?

Let $S(t)$ denote the survival function, i.e. $S(t)=P(T>t)$ is the probability 
the event of interest happens after time $t$. For the TRACE data is the 
probability a death occurs $t$ years after a heart attack.

The median survival time is the time at which $S(t)=0.5$. In other words, the 
median survival time is the time at which half of the population has already had
the event of interest occur.

## Estimating the median survival time by hand
The idea for estimating the median survival time by hand: find the time at which 
the estimated survival function is $0.5$.

```{r, fig.width = 7, fig.height = 5, echo=FALSE}
plot(survfit.trace, conf.int = TRUE, xlab = "Time (Years)", 
     ylab = "Survival Probability")
segments(x0 = 0, y0 = 0.5, x1 = 6.52, y1 = 0.5, col = "red")
```

## Estimating the median survival time by hand
The idea for estimating the median survival time by hand: find the time at which 
the estimated survival function is $0.5$.

```{r, fig.width = 7, fig.height = 5, echo=FALSE}
plot(survfit.trace, conf.int = TRUE, xlab = "Time (Years)", 
     ylab = "Survival Probability")
segments(x0 = 0, y0 = 0.5, x1 = 6.52, y1 = 0.5, col = "red")
segments(x0 = 6.52, y0 = 0, x1 = 6.52, y1 = 0.5, col = "red")
```

## Estimating the median survival time by hand
From the previous plot it looks like the median is somewhere between 6 and 7. We 
can use the output from \texttt{survfit} to make this estimate more precise. In 
particular, our estimate of the median is the smallest time such that the 
estimated survival function is less than $0.5$. Thus our estimate here is 
$6.53$.

\footnotesize
```{r}
summary(survfit.trace, times = seq(from = 6.50, to = 6.55, by = 0.01))
```
\normalsize

## Constructing a confidence interval for the median survival time by hand
The idea for constructing a confidence interval for the median survival time by 
hand is similar to estimating the median: find the times at which 
the upper and lower bounds of the confidence interval for the survival function 
are $0.5$.

```{r, fig.width = 7, fig.height = 5, echo=FALSE}
plot(survfit.trace, conf.int = TRUE, xlab = "Time (Years)", 
     ylab = "Survival Probability")
segments(x0 = 0, y0 = 0.5, x1 = 5.36, y1 = 0.5, col = "red")
segments(x0 = 5.36, y0 = 0, x1 = 5.36, y1 = 0.5, col = "red")
segments(x0 = 0, y0 = 0.5, x1 = 7.38, y1 = 0.5, col = "blue")
segments(x0 = 7.38, y0 = 0, x1 = 7.38, y1 = 0.5, col = "blue")
```

## Constructing a confidence interval for the median survival time by hand
From the previous plot it looks like lower bound of the confidence interval is 
between 5 and 6, and the upper bound is between 7 and 8. We can use the output 
from \texttt{survfit} to make this estimate more precise. In 
particular, our confidence interval lower bound is the smallest time such that 
the lower bound  is less than $0.5$. Thus our confidence interval lower bound 
here is $5.36$.

\footnotesize
```{r}
summary(survfit.trace, times = seq(from = 5.30, to = 5.40, by = 0.01))
```
\normalsize

## Constructing a confidence interval for the median survival time by hand
Our confidence interval upper bound is the smallest time such that 
the upper bound is less than $0.5$. Thus our confidence interval lower bound 
here is $7.38$.

\footnotesize
```{r}
summary(survfit.trace, times = seq(from = 7.35, to = 7.40, by = 0.01))
```
\normalsize

## Estimating the median survival time in \texttt{R}
We've done things the hard way, now let's let \texttt{R} do the rest. Note that 
the estimate of the median we constructed by hand is slightly off, as there was
an event in the data set between times $6.52$ and $6.53$ (we were just looking
at the survival function in $0.01$ year increments).

\footnotesize
```{r}
survfit.trace
```
\normalsize


## What is the median residual survival time?

Let $S(t| t_0)$ denote the survival function conditional on survival at time 
$t_0$, i.e. $S(t| t_0)=P(T>t|T>t_0)$ is the probability 
the event of interest happens after time $t$, given that the event hasn't 
already occurred at time $t_0$. For the TRACE data is the 
probability a death occurs $t$ years after a heart attack, given death hadn't 
already occurred at time $t_0$.

The median residual survival time is $t-t_0$, where $t$ is the time at which 
$S(t|t_0)=0.5$. In other words, after subsetting our population to those that 
haven't had an event at time $t_0$, the median residual survival time is how 
long after $t_0$ it takes for half of this sub-population to have the event.

## Estimating the median residual survival time by hand

Suppose we're interested in the median residual survival time, where $t_0=0.25$ 
(roughly one month). To estimate this by hand, we can follow slide 107 of the 
chapter 2 slides. In particular, if we can find the smallest time $t^*$ such that 
$S(t^*)\leq 0.5S(t_0)$, the median residual survival time is $t^*-t_0$.

From the code below, our estimate of $S(t_0)$ is $0.865$. 

\footnotesize
```{r}
summary(survfit.trace, times = 0.25)
```
\normalsize

## Estimating the median residual survival time by hand
Th previous slide tells us we want to find the smallest time $t^*$ such that 
$S(t^*)\leq 0.5\times 0.865=0.4325$.

From the code below, this time is $t^*=7.83$, and thus our estimate of the 
median residual survival time is $t^*-t_0=7.83-0.25=7.58$ years.

\footnotesize
```{r}
summary(survfit.trace, times = seq(from = 7.80, to=7.85, by = 0.01))
```
\normalsize

## Estimating the median residual survival time in \texttt{R}
We've done things the hard way, now let's let \texttt{R} do the rest. Using the 
\texttt{getmedianres} function, we arrive at an estimate of $7.575$, and a 
confidence interval of $[6.559, 7.575]$ (note that the upper bound is equal to 
the estimate, a consequence of the longest observed time corresponding to a 
censored event).

```{r, echo = FALSE}
set.seed(42)
```
\footnotesize
```{r}
getmedianres(s.trace, times = 0.25, confint = TRUE)
```
\normalsize

## Q & A time!

Any questions about the material covered today?

Any questions about anything else related to the course?
