---
title: 'Discussion 2: Fitting Parametric Models in R'
author: "Serge Aleshin-Guendel"
date: "1/18/2021"
fontsize: 10pt
output:
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

I'll try to stick to the same outline as Marlena:

- Review concepts from lecture for 35 minutes (please ask questions!)
- 15 minutes for open Q&A

Let me know what's useful for y'all! More time reviewing concepts, more time for 
Q&A, going over homework solutions (after homeworks are graded), etc.

## Learning Objectives

- How to fit and plot a Kaplan-Meier curve
- How to fit parametric models to survival data
- How to plot survival functions from parametric model fits 
- How to compare fits of nested parametric models
- How to estimate various measures from parametric models

## Slides

- Slides are available on Canvas
- RMarkdown can be checked out via GitHub

```{sh, eval=F}
git clone
 https://github.com/mbannick/survival-discussion-section.git
```


## Review: TRACE Data

TRACE study of myocardial infarction (Jensen et al. 1997)

Participants are patients with myocardial infarction admitted to the coronary care unit at Glostrup County Hospital in Denmark between 1977 and 1988.

Primary objective was to evaluate the relationship  between various risk factors and survival after myocardial infarction.

We utilize here a subset of 500 patients available in the \texttt{timereg} package in \texttt{R}.

## Review: TRACE Data

Patient characteristics in this subsample include:

- 259 deaths due to myocardial infarction and 5 due to another cause; all others were alive at the end of the study (prop. censoring: 47.2\%).
- 338 women + 162 men = 500 patients.
- Average age at myocardial infarction: 67.5 (range: 28.1--92.1).
- Comorbidities present:
     - 261 patients had congestive heart failure,
     - 29 patients had ventricular fibrillation,
     - 47 patients had diabetes.

## Review: TRACE Data

Relevant variables in dataset:

- \texttt{time}: time until death or censoring
- \texttt{status}: description of terminating event
     - 0 = alive at end of study
     - 9 = death due to myocardial infarction
     - 7 = death due to other causes
     
- \texttt{wmi}: wall motion index (ultrasound-based measure of heart function)
     - $<$ 2 = akinesis, hypokinesis and dyskinesis 
     - $>$ 2 = hyperkinesis
     
- \texttt{chf}: indicator of congestive heart failure
- \texttt{age}: age of patient (in years)
- \texttt{sex}: sex (\ 0 = female, 1\ =\ male\ )
- \texttt{diabetes}: indicator of diabetes
- \texttt{vf}: indicator of ventricular fibrillation

## Reading in the TRACE data

Note that in order to source \texttt{fitparametric.R} as we do below, it needs 
to be located in your working directory

```{r, warning = FALSE}
# Load relevant packages 
library(foreign)
library(survival)
library(flexsurv)

# Load in fitparametric function
source("fitparametric.R")

# Read in TRACE data
trace <- read.dta("traceST11.dta")
```

## Reading in the TRACE data

```{r, warning = FALSE}
head(trace)
```

## Censoring in the TRACE data set

Following Jon's course slides, we will treat both end of study and death due to 
causes other than myocardial infarction as censoring events

```{r}
# Create a Surv object using the TRACE data
s.trace <- with(trace, Surv(time, status == 9)) 
```

## Estimating the survival function nonparametrically

Before fitting parametric models, estimate the survival function 
nonparametrically, and plot the estimate:

\footnotesize
```{r, eval = FALSE}
survfit.trace <- survfit(s.trace ~ 1, data = trace, conf.type = "log-log")
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
```
\normalsize

## Estimating the survival function nonparametrically

```{r, fig.width = 10, fig.height = 8, echo=FALSE}
survfit.trace <- survfit(s.trace ~ 1, data = trace, conf.type = "log-log")
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
```


## Fitting an exponential model

Fit an exponential model to the TRACE data

\footnotesize
```{r}
fit_exp <- fitparametric(s.trace, dist = "exp")
```
\normalsize

## Plotting exponential model fit

Here we plot the estimated survival function based on the exponential model fit,
and compare it to the nonparametric estimate
\footnotesize
```{r, eval = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
legend("topright", c("nonparametric estimator", "exponential"), 
       col = c("black", "darkgreen"), lty = c("solid", "solid"), 
       lwd = rep(3, 4), cex = 0.9)
```
\normalsize

## Plotting exponential model fit

```{r, fig.width = 10, fig.height = 8, echo = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
legend("topright", c("nonparametric estimator", "exponential"), 
       col = c("black", "darkgreen"), lty = c("solid", "solid"), 
       lwd = rep(3, 4), cex = 0.9)
```

## Fitting a Weibull model

Fit an Weibull model to the TRACE data

\footnotesize
```{r}
fit_weibull <- fitparametric(s.trace, dist = "weibull")
```
\normalsize


## Plotting Weibull model fit

Here we plot the estimated survival function based on the Weibull model fit,
and compare it to the nonparametric estimate and the exponential estimate
\footnotesize
```{r, eval = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
lines(fit_weibull$fit, ci = FALSE, col = "orange", lty = "dashed", lwd = 2)
legend("topright", c("nonparametric estimator", "exponential", "weibull"), 
       col = c("black", "darkgreen", "orange"), 
       lty = c("solid", "solid", "dashed"), lwd = rep(3, 4), cex = 0.9)
```
\normalsize

## Plotting Weibull model fit

```{r, fig.width = 10, fig.height = 8, echo = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
lines(fit_weibull$fit, ci = FALSE, col = "orange", lty = "dashed", lwd = 2)
legend("topright", c("nonparametric estimator", "exponential", "weibull"), 
       col = c("black", "darkgreen", "orange"),
       lty = c("solid", "solid", "dashed"), lwd = rep(3, 4), cex = 0.9)
```

## Does the Weibull model provide a better fit than the exponential?

Suppose we wanted to test the hypothesis that the exponential model is an 
appropriate simplification of the Weibull model. Recall that in terms of the
parameterization used in the \texttt{fitparametric} function, the exponential
is a special case of the Weibull where the parameter $p=1$.

Thus we could rephrase the hypothesis as: is $p=1$?

## Does the Weibull model provide a better fit than the exponential?

\scriptsize
```{r}
fit_weibull <- fitparametric(s.trace, dist = "weibull")
```
\normalsize

A $95\%$ confidence interval for $p$ is $[0.49, 0.62]$, which doesn't 
include $p=1$. Thus we reject the hypothesis that the exponential model is an 
appropriate simplification of the Weibull model at the $\alpha=0.05$ 
significance level.

This is an example of a Wald test (see slide 20 of Chapter 2).

## Does the Weibull model provide a better fit than the exponential?

Alternatively, we could perform a likelihood ratio test (see slide 20 of Chapter
2). The statistic for the likelihood ratio test is 
$$\hat{T}_{LRT}=-2[\log(L_n(\theta_0)) - \log(L_n(\hat{\theta}))],$$ where:

- $\log(L_n(\hat{\theta}))$ is estimated log-likelihood under the full (Weibull)
model
- $\log(L_n(\theta_0))$ is estimated log likelihood under the reduced 
(exponential) model

\footnotesize
```{r}
T_LRT <- -2 * (fit_exp$loglik - fit_weibull$loglik)
T_LRT
```
\normalsize

## Does the Weibull model provide a better fit than the exponential?

For large $n$, under the null hypothesis $\hat{T}_{LRT}$ is approximately
chi-squared distributed, with degrees of freedom equal to $1$ (the difference in 
dimensionality for the exponential and Weibull models). A p-value for our 
likelihood ratio statistic can then be calculated as:
\footnotesize
```{r}
1 - pchisq(T_LRT, df = 1)
```
\normalsize

Note that the p-value is not exactly $0$, but numerically $0$. Again we reject 
the hypothesis that the exponential model is an appropriate simplification of 
the Weibull model at the $\alpha=0.05$ significance level.

## Fitting a generalized gamma model

Fit an generalized gamma  model to the TRACE data

\footnotesize
```{r}
fit_gengamma <- fitparametric(s.trace, dist = "gengamma")
```
\normalsize


## Plotting generalized gamma model fit

Here we plot the estimated survival function based on the generalized gamma  model fit,
and compare it to the nonparametric estimate, the exponential estimate, and the
Weibull estimate

\footnotesize
```{r, eval = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
lines(fit_weibull$fit, ci = FALSE, col = "orange", lty = "dashed", lwd = 2)
lines(fit_gengamma$fit, ci = FALSE, col = "blue", lty = "dotted", lwd = 2)
legend("topright", 
       c("nonparametric estimator", "exponential", "weibull", "gengamma"), 
       col = c("black", "darkgreen", "orange", "blue"),
       lty = c("solid", "solid", "dashed", "dotted"), 
       lwd = rep(3, 4), cex = 0.9)
```
\normalsize

## Plotting generalized gamma model fit

```{r, fig.width = 10, fig.height = 8, echo = FALSE}
plot(survfit.trace, conf.int = FALSE, xlab = "Time (in years)",
     ylab = "Survival probability", col = "black", lty = "solid", lwd = 2)
lines(fit_exp$fit, ci = FALSE, col = "darkgreen", lty = "solid", lwd = 2) 
lines(fit_weibull$fit, ci = FALSE, col = "orange", lty = "dashed", lwd = 2)
lines(fit_gengamma$fit, ci = FALSE, col = "blue", lty = "dotted", lwd = 2)
legend("topright", 
       c("nonparametric estimator", "exponential", "weibull", "gengamma"), 
       col = c("black", "darkgreen", "orange", "blue"),
       lty = c("solid", "solid", "dashed", "dotted"), 
       lwd = rep(3, 4), cex = 0.9)
```

## Does the generalized gamma model provide a better fit than the Weibull?

Suppose we wanted to test the hypothesis that the Weibull model is an 
appropriate simplification of the generalized gamma model. The parameterization
of the generalized gamma distribution in the \texttt{fitparametric} function 
does not lend itself to a Wald test this time, so we'll jump straight to the
likelihood ratio test.

## Does the generalized gamma model provide a better fit than the Weibull?

Just as before, the statistic for the likelihood ratio test is $$\hat{T}_{LRT}=-2[\log(L_n(\theta_0)) - \log(L_n(\hat{\theta}))],$$ where:

- $\log(L_n(\hat{\theta}))$ is estimated log-likelihood under the full 
(generalized gamma) model
- $\log(L_n(\theta_0))$ is estimated log likelihood under the reduced (Weibull) 
model

\footnotesize
```{r}
T_LRT <- -2 * (fit_weibull$loglik - fit_gengamma$loglik)
T_LRT
```
\normalsize

## Does the generalized gamma model provide a better fit than the Weibull?

For large $n$, under the null hypothesis $\hat{T}_{LRT}$ is approximately
chi-squared distributed, with degrees of freedom equal to $1$ (the difference in 
dimensionality for the Weibull and generalized gamma models). A p-value for our 
likelihood ratio statistic can then be calculated as:
\footnotesize
```{r}
1 - pchisq(T_LRT, df = 1)
```
\normalsize

We reject the hypothesis that the Weibull model is an appropriate simplification 
of the generalized gamma model at the $\alpha=0.05$ significance level.

## Estimating median time to death with parametric model

Suppose we want to estimate the median time to death using the generalized gamma
model. Can be done just by modifying the arguments to \texttt{fitparametric}:

\tiny
```{r}
library(numDeriv)
fit_gengamma <- fitparametric(s.trace, dist = "gengamma", feature = "quantile")
```
\normalsize

## Estimating survival function with parametric model

Suppose we want to estimate the probability that death doesn't occur in the 
first two years using the generalized gamma model. This can be written as 
$P(T>2)=S(2)$, i.e. the survivor function at two years. Can be done just by 
modifying the arguments to \texttt{fitparametric}:

\tiny
```{r}
fit_gengamma <- fitparametric(s.trace, dist = "gengamma", feature = "survival", 
                              t = 2)
```
\normalsize

## Estimating conditional survival function with parametric model

Suppose we want to estimate the probability that death doesn't occur in the 
first two years, given death hasn't occurred in the first year, using the 
generalized gamma model. This can be written as 
$P(T>2|T>1)=S(2)/S(1)$, i.e. the survivor function at two years, conditioned on 
survival at year one. Can be done just by modifying the arguments to 
\texttt{fitparametric}:

\tiny
```{r}
fit_gengamma <- fitparametric(s.trace, dist = "gengamma", 
                              feature ="condsurvival", t = 2, t0 = 1)
```
\normalsize

## Does the surival distribution differ between to sub-populations?

Suppose we want to determine whether or not the time to death distribution
differs between those with and without diabetes. Let's assume we're fitting 
separate exponential models to each sub-population, i.e. we're assuming
$T\sim\text{Exp}(\lambda_1)$ for those without diabetes and 
$T\sim\text{Exp}(\lambda_2)$ for those with diabetes. The time to death 
distribution differs between populations when $\lambda_1\neq\lambda_2$, or 
equivalently when $\delta:=\lambda_1-\lambda_2\neq 0$. 

## Does the surival distribution differ between to sub-populations?

Fit the separate exponential models

\footnotesize
```{r}
fit_exp_diabetes_1 <- flexsurvreg(Surv(time, status == 9) ~ 1, 
                                  data = trace[trace$diabetes == 0, ],
                                  dist = "exponential")
fit_exp_diabetes_2 <- flexsurvreg(Surv(time, status == 9) ~ 1, 
                                  data = trace[trace$diabetes == 1, ],
                                  dist = "exponential")
fit_exp_diabetes_1$res
fit_exp_diabetes_2$res
```
\normalsize

## Does the surival distribution differ between to sub-populations?

We have an estimate of $\delta$, i.e. 
$$\hat{\delta}=\hat{\lambda}_1 - \hat{\lambda}_2=0.1108693-0.1802261= -0.06935684.$$
If we can derive the standard error of $\hat{\delta}$, $SE(\hat{\delta})$, we 
can construct a Wald statistic
$$\hat{T}_{W}=\frac{|\hat{\delta}-\delta_0|}{SE(\hat{\delta})},$$
where $\delta_0=0$ under the null hypothesis that the time to death distribution
doesn't differ between those with or without diabetes.

\footnotesize
```{r}
delta <- fit_exp_diabetes_1$res[1] - fit_exp_diabetes_2$res[1]
delta
```
\normalsize


## Does the surival distribution differ between to sub-populations?
Assuming the two sub-populations are independent, the standard error of 
$\hat{\delta}$ is a function of the standard errors of $\hat{\lambda}_1$ and
$\hat{\lambda}_2$:
$$SE(\hat{\delta})=\sqrt{SE(\hat{\lambda}_1)^2 + SE(\hat{\lambda}_2)^2}.$$
\footnotesize
```{r}
delta_se <- sqrt(fit_exp_diabetes_1$res[4] ^ 2 + 
                     fit_exp_diabetes_2$res[4] ^ 2)
T_W <- abs(delta) / delta_se
T_W
```
\normalsize

## Does the surival distribution differ between to sub-populations?

For large $n$, the Wald statistic is normally distributed. A two-sided p-value
can then be calculated as follows:

\footnotesize
```{r}
2 * pnorm(-T_W) 
```
\normalsize

Thus at the $\alpha=0.05$ level, we reject the null hypothesis that the time to 
death distribution doesn't differ between those with or without diabetes.


## Q & A time!

Any questions about the material covered today?

Any questions about anything else related to the course?