---
title: 'Discussion 1: Censoring, Truncation, and R'
subtitle: 'BIOST 537'
author: "Marlena Bannick"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

The outline for these discussion sections (led by me) will generally be:

- Review concepts from lecture for 35 minutes (please ask questions!)
- 15 minutes for open Q&A

## Learning Objectives

- Understand how censoring arises in survival data
- Understand how truncation arises in survival data
- Recognize the difference between censoring and truncation
- Learn how to create survival data sets and survival plots in R

## Slides

- Slides are available on Canvas
- RMarkdown can be checked out via GitHub

```{sh, eval=F}
git clone
 https://github.com/mbannick/survival-discussion-section.git
```

## Poll Everywhere

We will use Poll Everywhere to take some quick anonymous polls
to stay engaged!

- In your browser on your phone or computer, navigate to PollEv.com/mnorwood
- You will be prompted to do a UW single sign on with your UW login
- Alternatively, text MNORWOOD to 22333 to join the session

## First Poll

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## First Poll

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/DKjjdclH3Mwn4rtYLTfcf?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Survival Data

- Survival data consists of information about time to an event.
- In order to create survival data, you need to define
  - Initiating event
  - Terminating event
  - Time scale

## Review: Survival Data

What makes survival data especially tricky is that we commonly see:

- Censoring
- Truncation

## Review: Censoring

Data can be left censored, right censored, or interval censored. In this course, we mainly focus on *right censored* data. It is incredibly common in survival data.

- **Right censoring**: When all you know about the event time is that it is beyond some time point (e.g. you will not observe the event for some individuals)
- **Left censoring**: When all you know about the event time is that it is before some time point (e.g. some have already had the event when they enter the study)
- **Interval censoring**: When all you know about the event time is that it is in between two time points (e.g. testing negative at one visit and positive at the next)

## Review: Censoring

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Censoring

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/Ur42wscWUZSu1TKv0wobZ?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Non-informative Censoring

There are techniques in survival analysis to account for censoring, such that the estimates of survival (through Kaplan-Meier, more later) or parameter estimation, are unbiased.

However, these methods require that the censoring is *non-informative*.

- **From lecture**: An individual who is censored is no more or less likely to experience the event in the immediate future than an individual who is not censored.

In other words, if an individual is censored, this should not give you any hints about whether or not they will experience the event sooner or later than anyone else.

## Review: Non-informative censoring

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Non-informative Censoring

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/FSmQXYygaceaFIxv8cEyq?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Truncation

Truncation occurs when an individual's time to event excludes them from being included in the sample.

This can happen when the time to event is too small, e.g. the terminating event happened before the study could enroll someone (left truncation).

This can happen when the time to event is too large, e.g. a retrospective study selects only individuals for which the terminating event has happened already (right truncation).

## Review: Truncation

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Truncation

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/DFoOuA31D1t6Brzlf6Gng?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Truncation versus Censoring

The fundamental difference between truncation and censoring is in who is included in the study.

- **Truncation induces selection bias**: because of their event time, individuals didn't even make it into the study.
- **Censoring is incomplete information**: for the individuals that did make it into the study, we know something about their event time (e.g. after some point), but don't know it precisely.

## Survival Packages

A package that includes very handy functions for dealing with survival
data is `survival`.

```{r}
library(survival)
```

## Survival Data

```{r}
data("diabetic")
head(diabetic)
```
## Survival Data

The "diabetic" dataset from package `survival`

- `id` subject id
- `laser` laser type: xenon or argon
- `age` age at diagnosis
- `eye` a factor with levels of left right
- `trt` treatment: 0 = no treatment, 1= laser
- `risk` risk group of 6-12
- `time` time to event or last follow-up
- `status` status of 0 = censored or 1 = visual loss

## Survival Objects

- `time` is the recorded amount of time from the event initiation until either the event happens or some censoring occurs
- `event` is an indicator for whether or not the observation was censored

```{r}
surv <- Surv(time=diabetic$time, event=diabetic$status)
head(surv)
```

## Survival Objects

```{r}
plot(surv)
```

## Survival Data

Many of the data sets that we will be working with have been pre-processed,
so that we can easily use the `Surv` function.

What happens if we encounter data in the wild?

- May have a "start time" (could be in datetime format like "2020-01-01")
- May have an "end time" but lots of missing data for the end times

## Survival Data

```{r, include=F}
library(lubridate)
library(data.table)
```

```{r}
set.seed(10)
begin <- dmy("20/10/2020")
today <- dmy("12/01/2021")

initiate <- begin + sample(0:10, size=100, replace=TRUE)
terminate <- initiate + sample(0:200, size=100, replace=TRUE)
terminate[terminate >= today] <- NA

df <- data.table(initiate=initiate, terminate=terminate)
head(df)
```

## Survival Data

How would we get `df` into a form that we could use with `Surv`?

```{r}
df[, time := as.numeric(terminate - initiate)]
df[, event := !is.na(time)]
df[is.na(time), time := as.numeric(today - initiate)]

head(df)
```

## Survival Data

```{r}
surv <- Surv(time=df$time, event=df$event)
plot(surv)
```
