---
title: 'Discussion 1: Censoring, Truncation, and R'
author: "Marlena Bannick"
output:
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

The outline for these discussion sections (led by me) will generally be:

- Review concepts from lecture for 35 minutes (please ask questions!)
- 15 minutes for open Q&A

## Learning Objectives

- Understand how censoring arises in survival data
- Understand how truncation arises in survival data
- Recognize the difference between censoring and truncation
- Learn how to create survival data sets and survival plots in R

## Poll Everywhere

We will use Poll Everywhere to take some quick anonymous polls
to stay engaged!

- In your browser on your phone or computer, navigate to PollEv.com/mnorwood
- You will be prompted to do a UW single sign on with your UW login
- Alternatively, text MNORWOOD to 22333 to join the session

## First Poll

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## First Poll

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/6fxdePiOwUhVB9NGmMjjS?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Slides

- Slides are available on Canvas
- RMarkdown can be checked out via GitHub

## Review: Survival Data

- Survival data consists of information about time to an event.
- In order to create survival data, you need to define
  - Initiating event
  - Terminating event
  - Time scale

## Review: Survival Data

What makes survival data especially tricky is that we commonly see:

- Censoring
- Truncation

## Review: Censoring

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Censoring

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/Ur42wscWUZSu1TKv0wobZ?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Non-informative Censoring

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Non-informative Censoring

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/FSmQXYygaceaFIxv8cEyq?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Review: Truncation

<iframe src="https://pollev-embeds.com/mnorwood" width="800px" height="600px"></iframe>

## Review: Truncation

<iframe src="https://embed.polleverywhere.com/multiple_choice_polls/DFoOuA31D1t6Brzlf6Gng?controls=none&short_poll=true" width="800px" height="600px"></iframe>

## Survival Packages

A package that includes very handy functions for dealing with survival
data is `survival`.

```{r}
library(survival)
```

## Survival Data

```{r}
data("diabetic")
head(diabetic)
```
## Survival Data

The "diabetic" dataset from 

- `id` subject id
- `laser` laser type: xenon or argon
- `age` age at diagnosis
- `eye` a factor with levels of left right
- `trt` treatment: 0 = no treatment, 1= laser
- `risk` risk group of 6-12
- `time` time to event or last follow-up
- `status` status of 0= censored or 1 = visual loss

## Survival Objects

- `time` is the recorded amount of time from the event initiation until either the event happens or some censoring occurs
- `event` is an indicator for whether or not the observation was censored

```{r}
surv <- Surv(time=diabetic$time, event=diabetic$status)
```

## Survival Data

Many of the data sets that we will be working with have been pre-processed,
so that we can easily use the `Surv` function.

What happens if we occur data in the wild?

- May have a "start time" (could be in datetime format like "2020-01-01")
- May have an "end time" but lots of missing data for the end times

## Survival Data

```{r, include=F}
library(lubridate)
library(data.table)
```

```{r}
set.seed(10)
begin <- dmy("20/10/2020")
today <- dmy("12/01/2021")

initiate <- begin + sample(0:10, size=100, replace=TRUE)
terminate <- initiate + sample(0:200, size=100, replace=TRUE)
terminate[terminate >= today] <- NA

df <- data.table(initiate=initiate, terminate=terminate)
head(df)
```

## Survival Data

How would we get `df` into a form that we could use with `Surv`?

```{r}
df[, time := as.numeric(terminate - initiate)]
df[, event := !is.na(time)]
df[is.na(time), time := as.numeric(today - initiate)]

head(df)
```

## Survival Data

```{r}
surv <- Surv(time=df$time, event=df$event)
plot(surv)
```
